Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex

% Finally, we’ll look at some papers that discuss implementing sustainability in scientific software.

% Blanton and Lenhardt from the Renaissance Computing Institute discuss these issues from a user perspective.

% The authors focus on a point that has been brought up many times in this context:
% There is a tension between writing code that is good enough just to “get it done,” i.e. to publish a paper about scientific results obtained using software, 
% and “getting it right,” that is, developing software that is comprehensible to future users and reviewers.
% Just because the elevator panel works like this doesn’t mean it’s sustainable for the long run.
% We don’t have a way to validate that the software used in a paper is actually done right.
% The best way to get software designed correctly is to make sure best practices are considered from the start.

% The authors highlight two models for sustainable software, at different extremes:
% One is what they call “co-funding”:
% In these projects, usually large, multi-year collaborations, 
% there is equal emphasis on both the science and the software development. 
% Both are planned into the project from inception. 
% In the life sciences, the iPlant Collaborative, Galaxy Project, and Qiime are good examples of these sorts of large, well-designed projects.

% At the other extreme, they discuss “software carpentry”: 
% in this model, it’s assumed that the scientists themselves will write and maintain their code.
% Groups like Software Carpentry and ROpenSci assume that 
% scientists won’t have access to dedicated software engineering, 
% so they try to give them tools to use best practices in their own software development.

% There might be a middle ground here: a way to get the engineering expertise that large co-funded projects have to individual scientist-developers.
% Hilmar Lapp of NESCent and I discuss one such possibility in our paper, Software Engineering as Instrumentation for the Long Tail of Scientific Software.

% What do we mean when we refer to the “long tail” of scientific software?
% Think of the distribution of resources in scientific software. Most are focused on big projects with lots of community buy-in and funding. But a lot of scientific software exists away from this model.
% For example, scientific software can be used long after the original developer has moved on or the funding runs out.
% Look at MacClade: it was originally released in 1986 and last updated in 2005,
% but it was still cited over 400 times in 2013!
% The scientists who developed it have a newer package, Mesquite, that was meant to replace MacClade, but they haven’t had sufficient time or resources to maintain either package fully, let alone both of them.

% Another dimension of the long tail can also be found in my particular research domain. 
% In the field of phylogenetics, we have a lot of programs that implement different computational methods in slightly different ways.
% Here, Joe Felsenstein has listed some (but not anywhere near all) phylogenetics packages available online.
% Most of these programs are developed by academic scientists…
% They generally have limited training in software engineering
% Limited time or career incentive to improve software
% Limited funding

% So, to summarize a bit:
% Making sustainable software means we have to pay attention to many facets of software design, like APIs, security, user experience, testing, etc.
% A single project that requires one full­-time software engineer may actually require fractions of different kinds of engineers. 
% But long-tail projects can’t even fund one FTE, let alone one that can address all these facets.

% Then we have to consider that the users of scientific software are scientists, 
% so the developers need to understand the users and the science.
% This is the idea of a “t-skilled” person: 
% one who is both well-versed in a scientific domain 
% and deeply experienced in one or more facets of software engineering. 
% These people are pretty rare in the first place and difficult to retain in academia, because the academic career structure doesn’t incentivize this.

% We should look at software engineering as an expensive resource, but one that needs to be accessible to scientists at all levels.
% Think of it as analogous to DNA sequencing:
% Sequencers used to be something that individual labs and institutions had to 
% buy, maintain, and operate themselves, 
% so only highly-funded operations had them and probably didn’t use them to their full capacity even when they had one.
% But now, core facilities provide the instrumentation and service to labs of any size. 
% Anyone can pay a core facility to sequence their samples for them and provide quality control and bioinformatics advice as additional services.

% We propose that software engineering can be “instrumented” in a similar way. 
% Let’s create a nonprofit center for scientific software engineering.
% This center can hire these t-skilled personnel and provide access to them for projects at contracted cost.
% Because the center is focused on providing development services to scientific projects, it is not tied to the long-term success or failure of any individual project.
% It would emphasize the centrality of doing good science by making functional software tools as envisioned by scientists.

% So, to conclude…
% Implementing policies to encourage sustainability in scientific software 
% requires that many facets of good software design are addressed throughout the lifecycle of these projects. 
% But most of them aren’t addressed in the status quo.
% We’ve highlighted some of these facets today and suggested some possible solutions.
% Large projects can afford to hire software engineers with the expertise to implement these facets correctly.
% Grassroots developer groups can provide guidance to scientists about best practices in software development.
% We think there is a place for a software engineering center that can provide 
% both engineering expertise and guidance 
% with a contract-driven instrumentation model 
% to the scientific software in the long tail.

>>>>

API governance and cybersecurity measures appropriate for a project's
maturity are all but two facets of sustainability-oriented software
development. Others include user-centered interface design, test
engineering, dependency management, and deployment operations. Each of
these facets require specialized skills and training in software
engineering. How can the implementation of best practices and
guidelines along many or all of these different facets be facilitated
in scientific software development? Blanton and
Lenhardt~\cite{Blanton_WSSSPE} contrast two models. In one, the time
and personnel devoted to software engineering is ``co-funded'' with
the driving science project. This typically implies large multi-year
collaborative projects that to succeed require significant software
infrastructure to be built, and which thus have the funding to support
one or several software engineers. The sustainability of such projects
then depends on sustaining the funding. In the other model, domain
scientists also take the role of software developers, whether by
necessity such as funding limitations, or due to cross-disciplinary
professional interests. For this to result in sustainable software
products, the domain scientists need to be (or become) cross-trained
in software engineering standards and best practices.

In practice, there are example for both extremes of the spectrum. For
example, in the life sciences the iPlant Collaborative~\cite{iPlant},
the Galaxy Project~\cite{Galaxy}, and Qiime~\cite{Qiime} are large
multi-year projects with significant software infrastructure funding
and deliverables. Typically though these are the exception rather than
the rule, and particularly in the long tail of science the
scientist-developer predominates. For providing these with better
training in software engineering best practices, initiatives such as
Software Carpentry~\cite{SoftwareCarpentry} have been especially
valuable, and Software Carpentry also shown how such training could
scale.

However, there may also be a middle ground between the two
extremes. Huang and Lapp~\cite{Huang_WSSSPE} propose a Center of
Excellence model that leverages economies of scale to make software
engineering experts and their skills accessible to the long tail of
science. As Huang and Lapp discuss, this model could effectively turn
the utilization of software engineering expertise from a complex human
resource recruitment and management challenge, to buying time, when
and to the extent needed, on shared instrumentation. There is
precedence to using such a model to lower access barriers for long
tail science, in particular for new experimental technologies.  For
example, although the acquisition and operation of next-generation
high-throughput DNA sequencers requires substantial investments of
money, time, and expertise, the establishment of ``core facilities''
on many university campuses has made these technologies accessible to
a wide swath of scientists, with transformative results for science.

One of the important conclusions from this group of papers is that
creating sustainable software requires paying attention not to one or
two, but to several different facets of software engineering, each
with corresponding best practices and standards of excellence. Even if
a science project requires and has funding for a full software
engineering FTE, what the project really needs could be fractions of
different engineers with different specialty training. The vast
majority of long tail science projects lacks the funding for even one
full software engineer, let alone one who combines expertise in all of
the applicable facets of engineering. Some scientists in the long tail
will have the professional interests to cross-train enough in software
engineering to be successful with creating sustainable software, but
it is unrealistic to expect this expertise and interest of all or even
the majority. This is where a software engineering center of
excellence could provide a critical resource by enabling scientists in
the long tail to complement their resources and expertise with facets
that are missing, but which, if applied at the right time, would
improve the chances of a software product to become sustainable. Such
complementary expertise also need not be restricted to software
engineering in the strict sense; for example, it could consist of
community building, leadership, and support for some period of
transition to sustainability.

In summary, implementing software sustainability practices on a
broader basis requires on the one hand the development of guidelines
and practices that are suitable for the exploratory research context
in which most scientific software is created, and on the other hand a
skilled workforce trained in a variety of software engineering facets
and community building. The capabilities afforded by such a workforce
need to be accessible not only to large projects with sufficient
funding to provide competitive employment, but also to the many
smaller projects in the long tail of science. Sufficiently
cross-training domain scientists could be one way to achieve this;
another, complementary approach is to instrument the necessary
capabilities so they can be shared.
